\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-27 paper/experiments.tex
Add second experiment results to paper

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is well suited for early dataset exploration.

\subsection{Data sets} \label{subsec:datasets}
\subsubsection{Center for Disease Control: Surveys}
We use survey data collected by the 
Center for Disease Control and Prevention (CDC) in the United States. We
experiment on a set of tables collected as part of the 2013-2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others. Our dataset is available for direct download on the CDC's organization site on 
kaggle.com, a popular data science website.

There are a total of 6 tables associated with the NHANES data set. We use three tables:

@@

\begin{table}
    \centering
    \begin{subtable}{0.5\textwidth}
        \centering
       \input{tables/cdc_demo}
        \caption{Demographics}
    \end{subtable}
    \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_labs}
        \caption{Laboratory Results}
    \end{subtable}
       \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_exams}
        \caption{Physical Results}
    \end{subtable}
    \caption{Missing value distribution within each table/attribute in CDC NHANES 2013-2014 data}
    \label{table:nhanes-description} 
\end{table}

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where many missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
   \centering
    \input{tables/fcc}
    \caption{Missing values in fCC Survey Data}
   \label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collect a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
to plan. They could reasonably be written by a user in the course of data analysis.

@@

Queries 1-4 are on the CDC data. In~\Cref{q1}, we calculate the average height for
users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
individuals with low, medium, and high incomes spectrum and above a certain weight. In~\Cref{q3}, 
we extract the maximum blood lead levels for children under the age of 6
years. In~\Cref{q4}, we calculate the average systolic blood pressure, by gender, for
subjects with a body mass index indicating obesity.

Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on fCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\Cref{fig:runtimes} shows a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\todo{change this to a single graph of X-times faster than base approach}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query ran for 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

@@

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

\begin{figure}
\begin{lstlisting}
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
    and 1 second, respectively.
}
\label{query-acs}
\end{figure}

>>>>

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

\subsection{Data sets} \label{subsec:datasets}
\subsubsection{Center for Disease Control: Surveys}
For our first set of experiments, we use survey data collected by the 
Center for Disease Control and Prevention (CDC) in the United States. In particular, we
experiment on a set of tables collected as part of the National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others. Our dataset corresponds to the 2013-2014 NHANES
and is available for direct download on the CDC's organization site on 
kaggle.com, a popular data science website.

There are a total of 6 tables associated with the NHANES data set. For purposes
of our experiments, we use three tables:

@@

\begin{table}
    \centering
    \begin{subtable}{0.5\textwidth}
        \centering
       \input{tables/cdc_demo}
        \caption{Demographics}
    \end{subtable}
    \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_labs}
        \caption{Laboratory Results}
    \end{subtable}
       \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_exams}
        \caption{Physical Results}
    \end{subtable}
    \caption{Missing values in CDC NHANES 2013-2014 data}
    \label{table:nhanes-description} 
\end{table}

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where plenty of missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
   \centering
    \input{tables/fcc}
    \caption{Missing values in FCC Survey Data}
   \label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
in that they could reasonably be written by a user in the course of data analysis, and
interesting to plan.

@@

Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we calculate the average height for
users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
individuals with low, medium, and high incomes spectrum and above a certain weight. In~\Cref{q3}, 
we extract the maximum blood lead levels for children under the age of 6
years. In~\Cref{q4}, we calculate the average systolic blood pressure, by gender, for
subjects with a body mass index indicating obesity.

Our second set of queries is on the FCC data (\Cref{fig:queries-fcc}). In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{query-acs}
 \caption{Queries on ACS data}
 \label{fig:query-acs}
 \end{subtable}

  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\Cref{fig:runtimes} provides a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{} imputation
    optimizing for quality ($\alpha=0$), and \ProjectName{} imputation optimizing for
    efficiency ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality. For
    relatively large queries (Query 9), runtime of imputation can be reduced from 6.5
    hours to 7 seconds. }
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. After a warm-up period, each query was run 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

@@

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(\Cref{fig:query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 355 minutes(\ref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.
